---
title: "HW1"
output: github_document
date: "2023-01-17"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
knitr::opts_chunk$set(warning = FALSE, message = FALSE) 

library(knitr)
library(caret)
library(rsample)
library(modelr)
library(tidyverse)
library(ggplot2)
library(readr)
```


## Question 1
### What is the best day of the week to fly, and does this change by airline?

```{r, include = FALSE}
airline = read_csv("ABIA.csv")
summary(airline)


```


```{r, echo = FALSE}
x = str_wrap("From this graph, we can see that the departure delay is fairly similar across all days of the week.  It is slightly worse Tuesday and Saturday, but nothing overly alarming.  It does seem that American Airlines has significantly higher departure delays on Saturdays.", 80)
ggplot(airline, aes(x = DayOfWeek, y = DepDelay, color = UniqueCarrier)) + geom_point() + xlab("Day of the Week") + ylab("Departure Delay") + ggtitle("Departure Delay by the Day of the Week (colored by Airline)") + labs(caption = x)


```


```{r, echo = FALSE}

library(dplyr)
y = str_wrap("Similarly to departure delays, we can see that the arrival delay is fairly similar across all days of the week.  It is slightly worse Tuesday and Saturday, but nothing overly apparent.  American Airlines also has significantly higher arrival delays on Saturdays than any other airline.", 80)
ggplot(airline, aes(x = DayOfWeek, y = ArrDelay, color = UniqueCarrier)) +geom_point() + xlab("Day of Week") + ylab("Arrival Delay") + ggtitle("Arrival Delay by Day of Week (Colored by Airline)") + labs(caption = y)


```

```{r, echo = FALSE}
avg_air = airline %>%
  group_by(DayOfWeek, UniqueCarrier) %>%
  summarise(dep_delay = mean(DepDelay), arr_delay = mean(ArrDelay), avg_delay = mean(arr_delay + dep_delay))

z = str_wrap("When we look at the average total delay, we can see that more airlines have delays on Tuesdays, but there is not much else we can reason from this graph", 80)

ggplot(avg_air, aes(x = DayOfWeek, y = avg_delay, color = UniqueCarrier)) + geom_point() + xlab("Day of the Week") + ylab("Average Total Delay (Including Arrival and Departure") + ggtitle("Average Delay by the Day of the Week (colored by Airline)") + labs(caption = z)



```

Overall, it seems as though Tuesdays and Saturdays experience slightly higher delays than average, and that American Airlines experiences larger delays on Saturdays. 

## Question 2




```{r, include = FALSE}

## read in data for 2

olympics = read_csv("olympics_top20.csv")

```

### Part a

The 95% of female heights across all athletics events is 183 centimeters.


```{r, include = FALSE}

##2a

oly_w_a = olympics %>%
  filter(sex == "F", sport == "Athletics")
quantile(oly_w_a$height, .95)
  

```

### Part b

Rowing Women's Coxed Fours had the highest standard deviation in height (10.865)

```{r, include = FALSE}
## 2b
oly_sd = olympics %>% 
  filter(sex == "F") %>%
  group_by(event) %>%
  summarise(sd_height = sd(height)) 

oly_sd[order(-oly_sd$sd_height),]


```
### Part c

In general, the average age of swimmers across genders has gradually increased over the years.  However, when the sport was first included in the olympics, it only was for males, and the average age dipped sharply when females were added to the sport. 

```{r, echo = FALSE}
olympic_swim = olympics %>%
  filter(sport == "Swimming")%>%
  group_by(sex, year) %>%
  summarise(average_age = mean(age))

ggplot(olympic_swim, aes(x = year, y = average_age, color = sex)) + geom_line()


```

## Question 3

### Information for trim = 350


```{r, echo = FALSE}
sclass = read_csv("sclass.csv")
sclass350 = sclass %>%
  filter(trim == "350")
sclass350_split =  initial_split(sclass350, prop=0.8)
sclass350_train = training(sclass350_split)
sclass350_test  = testing(sclass350_split)

```

```{r, include = FALSE}
set.seed(123)
## Attempt for loop
k_vec= c(2:332)
knnk = list()
output = list()
rmse_vec = c()

## for loop trim = 350

for (i in 2:length(k_vec)){
  knnk[[i]] = knnreg(price~mileage, data = sclass350_train, k =  k_vec[i])
output[i] = sclass350_test %>%
  mutate(Price_Pred = predict(knnk[[i]], sclass350_test))
 ##sclass350_test = sclass350_test %>%
 ## mutate(Price_pred = predict(knnk, sclass350_test))
rmse_vec[i] = modelr::rmse(knnk[[i]], sclass350_test)
}

k_vec
rmse_vec
plot(k_vec, rmse_vec)
output
my_data = data.frame(k_vec, rmse_vec)
my_data
min(my_data$rmse_vec)

```



```{r, echo = FALSE}
## 350 trim - k vs. RMSE plot 

plot(k_vec, rmse_vec)

```



```{r, include = FALSE}

## plot for k = 10 

knn10 = knnreg( price ~ mileage, data=sclass350_train, k=10)


## Predictions Knn10

sclass350pred_test = sclass350_test %>%
  mutate(Price_pred = predict(knn10, sclass350_test))

##RMSE KNN10
modelr::rmse(knn10, sclass350pred_test)

## Plot KNN10

p_test = ggplot(data = sclass350pred_test) + 
  geom_point(mapping = aes(x = mileage, y = price), alpha=0.2) + 
  ylim(7000, 20000)
p_test


```

```{r, echo = FALSE}
# now add the predictions
p_test + geom_line(aes(x = mileage, y = Price_pred), color='red', size=1.5)
```
### trim = 65

```{r, include = FALSE}

## data split for 65 AMG 
sclass
sclass65 = sclass %>%
  filter(trim == "65 AMG")
sclass65_split =  initial_split(sclass65, prop=0.8)
sclass65_train = training(sclass65_split)
sclass65_test  = testing(sclass65_split)

```



```{r, include = FALSE}
## for loop for 65 AMG

set.seed(123)
## Attempt for loop

k_vec65= c(2:17)
knnk65 = list()
output65 = list()
rmse_vec65 = c()

for (i in 2:length(k_vec65)){
  knnk65[[i]] = knnreg(price~mileage, data = sclass65_train, k =  k_vec65[i])
output65[i] = sclass65_test %>%
  mutate(Price_Pred = predict(knnk65[[i]], sclass65_test))
rmse_vec65[i] = modelr::rmse(knnk65[[i]], sclass65_test)
}

k_vec65
rmse_vec65
plot(k_vec65, rmse_vec65)

my_data65 = data.frame(k_vec65, rmse_vec65)
my_data65
min(my_data65$rmse_vec)

```
The minimum RMSE for the AMG 65 trim is 18150.36 at k = 13. 

### Plot for RMSE vs K, 65 AMG 

```{r, echo = FALSE}
plot(k_vec65, rmse_vec65)
```

### Plot for predictions based on mileage, including p-value

```{r, include = FALSE}

## Manual calculation for k = 13

knn13 = knnreg( price ~ mileage, data=sclass65_train, k=13)


## Predictions Knn13

sclass65pred_test = sclass65_test %>%
  mutate(Price_pred = predict(knn13, sclass65_test))

##RMSE KNN13
modelr::rmse(knn13, sclass65pred_test)

## Plot KNN13

p_test = ggplot(data = sclass65pred_test) + 
  geom_point(mapping = aes(x = mileage, y = price), alpha=0.2) 
p_test
```

```{r, echo = FALSE}
# now add the predictions
p_test + geom_line(aes(x = mileage, y = Price_pred), color='red', size=1.5)

```

The 65 AMG trim yields a larger optimal value for K.  I would predict that this is because there are less observations of the 65 AMG trim, so it is more difficult to find groupings with a smaller number of neighbors.  There are for more observations of the 350 trim, and so it takes a smaller number of neighbors to minimize the RMSE. 


```{r,include = FALSE}

## This code is the first attempt - rather than go by hand until I found the minimum I decided to use a for loop
## KNN 
##KNN2
knn2 = knnreg( price ~ mileage, data=sclass350_train, k=2)


## Predictions Knn2

sclass350_test = sclass350_test %>%
  mutate(Price_pred = predict(knn2, sclass350_test))

##RMSE KNN2
modelr::rmse(knn2, sclass350_test)

## KNN 
##KNN20
knn20 = knnreg( price ~ mileage, data=sclass350_train, k=20)


## Predictions Knn20

sclass350_test = sclass350_test %>%
  mutate(Price_pred = predict(knn20, sclass350_test))

##RMSE KNN2
modelr::rmse(knn20, sclass350_test)


## KNN 
##KNN50
knn50 = knnreg( price ~ mileage, data=sclass350_train, k=50)


## Predictions Knn2

sclass350_test = sclass350_test %>%
  mutate(Price_pred = predict(knn50, sclass350_test))

##RMSE KNN2
modelr::rmse(knn50, sclass350_test)

```


```{r ref.label=knitr::all_labels(), echo=TRUE, eval=FALSE}
```


